Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-06-03 docs/parasite_mortality_manuscript.tex
Incorporated all draft3 comments and added new figures.

\maketitle

\section*{Abstract}

Parasites can often have significant impacts on host populations through parasite-induced host mortality.  Detecting and quantifying the magnitude of parasite-
induced mortality, particularly for macroparasites where pathology is
linked to to the intensity of infection, can provide important insight into
parasite transmission and host population management. This is, however,
notoriously difficult in practice as the only data often available are measures of parasite intensities. In this study we provide a novel method for detecting and
quantifying parasite-induced mortality in wildlife populations from intensity
data and, using simulations, we show that our method is more reliable for
detecting and quantifying parasite-induced host mortality than the methods that are
currently available.  We also show that this method provides quantitative
assessments of parasite-induced mortality in empirical data that are consist
with previously published qualitative assessments. However, we stress that this
method, along with all methods for estimating parasite-induced mortality in
host populations, have a number critical assumptions that limit their
applicability in the real world.  We conclude that methods for estimating and
detecting parasite-induced mortality from intensity data should only be used as an
exploratory tool for informing more rigorous studies of parasite-induced host
mortality.

@@

Infectious agents can have major impacts on animal populations through changing
population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
of these infectious agents in wildlife is critical to understanding what
regulates host and parasite populations, making predictions about disease
transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine TB \citep{Cox2005}, and
rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of
disease, and does not account for the number of infectious agents present.
This method is sufficient for many bacterial and viral agents that reproduce within a
host, however for macroparasites pathology is linked to the intensity of infection and hosts cannot be simply categorized as infected and
uninfected \citep{AndersonandMay1979,Lafferty2002}.  Helminths exhibiting this intensity dependent pathology have
significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013},
wildlife survival \citep{Kirk2003, Logiudice2003}.  While it is generally assumed that some fraction of
wild host populations must succumb to parasitic infections, it is notoriously
difficult to actually quantify parasite-induced host mortality (PIHM) in wild
animal populations \citep{McCallum2000a}.

Ideally, parasite-induced host mortality is
quantified by experimentally infecting and tracking individual hosts in the
wild population; however, for logistical and ethical reasons this method is
rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and quantify the
relationship between infection intensity and host mortality \citep{Adjei1986}.

\cite{Crofton1971a} first proposed that PIHM could be identified by comparing the
observed parasite distribution in the host population to the distribution
predicted in the absence of parasite-induced mortality. We briefly introduce the Crofton Method here and provide a more detailed explanation of its implementation in \emph{Supplementary Material (SI)} 1. This method
assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts. Assuming mortality occurs only in these heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
fitting a negative binomial distribution to hosts with lower and lower parasite
loads, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data.

The Crofton Method may be able to detect the presence of PIHM, but it does not quantify the relationship between infection intensity and host survival
probability. \cite{Adjei1986} suggested that this relationship
could be calculated by first using the Crofton Method to estimate the pre-mortality parasite distribution and then using this distribution to calculate the
probability of host survival with increasing parasite intensity. To do this,
\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI} 2 for a technical description of the Adjei Method).
This method appeared to provide an estimate for the parasite intensity at which
50\% of hosts exhibit PIHM ($LD_{50}$), as well as the unmeasurable fraction of the
population that was lost (\emph{SI} 2). However, to implement this method the observed data must be
modified to fit the GLM framework and subjectively binned when mean
infection intensity is high or sample sizes are small.

After 30 years, and despite clear limitations \citep{McCallum2000a}, these
methods (particularly the Crofton Method) are still discussed among
parasitologists and are the primary techniques for examining population level
impacts of parasitism using parasite intensity data. In these methods, PIHM can
only be identified by visually examining plots and, with no clear decision
rule, it can be difficult to determine the significance of PIHM across
different host-parasite systems. The survival function given by the Adjei
Method offers one solution; however, this method requires manipulating the
original data and has never been tested.

Intensity data should only
be used to estimate parasite impacts on host populations if unbiased and accurate methods exist. In
this study, we first propose a novel method for detecting and quantifying PIHM. We
next use simulations to compare our method with the previous Adjei
Method to test the ability of both to (1) detect occurrence of PIHM and (2) estimate the
lethal parasite load ($LD_{50}$) and the associated survival function.  We then
apply both methods to real datasets previously used in PIHM analyses and
compare the results. Finally, we discuss the limitations of inferring PIHM
from intensity data and whether any method for inferring PIHM has a place
in quantitative parasitology.

@@

Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
require binning or data alteration, reduces the number of
parameters to be estimated, and  uses standard statistical techniques to determine PIHM significance. We provide Python code for implementing the Likelihood Method in \emph{SI} 4.

As with all previously proposed methods for estimating PIHM, the
Likelihood Method first assumes that prior to mortality the parasite distribution is described by a negative binomial $g(x; \mu_p, k_p)$, where $\mu_p$ and $k_p$ are the mean parasite intensity and aggregation before mortality, respectively (smaller $k_p$ indicates more aggregation).   Previous methods required calculating the total number of hosts before mortality ($N_p$) \citep{Crofton1971a,Adjei1986}, however this parameter is not needed in the Likelihood Method.

The Likelihood Method then assumes that the host survival function, which specifies the probability of a host surviving with $x$ parasites, follows the logistic curve given by

\begin{equation}
    h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}
With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$ by  first defining a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.  Using standard rules of conditional probability this distribution can be written as

@@

One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$. Therefore equation \ref{eq:concept} can be written as

\begin{equation}
    P(x | \text{survival}) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
    \label{eq:dist}
\end{equation}

Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset. The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.  All parameters are defined in Table \ref{tab:params}.

To estimate the significance of PIHM in a host-parasite system, a
likelihood ratio test is used in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by a negative binomial
distribution.  If PIHM is not significant in the system, the resulting likelihood
ratio statistic should approximately follow a $\chi^2$ distribution with two degrees of freedom.

@@

We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters. First, we created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$. Second, we chose values of $a$ and $b$ and calculated the probability of survival
for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, for each host, we drew a random number from a uniform distribution
between 0 and 1 and if the calculated host survival probability was less than this random
number, the host experienced parasite-induced mortality. The parasite distribution in these simulated surviving hosts is equivalent to the observed parasite distribution in a wild host population that has undergone parasite-induced host mortality.

We used these simulated pre-mortality and post-mortality datasets to test the
ability of both methods to correctly determine whether or not PIHM was
occurring when the parameters $N_p$, $\mu_p$ and $k_p$ were known.  Although
the parameters $N_p$, $\mu_p$, and $k_p$ are always unknown in real systems, a
method that fails under these ideal simulation conditions will certainly also
fail using less ideal, empirical data. In practice, for the Adjei Method, $N_p$, $\mu_p$,
and $k_p$ are estimated using the Crofton Method \citep{Adjei1986}, while $\mu_p$ and $k_p$ in
the Likelihood Method can be estimated jointly with $a$ and $b$ or via the
Crofton Method.

We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:question1}A).  For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.  We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500]. $N_p$ is not technically the sample size on which the methods are being
tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.\\

\noindent
\emph{Question 2: Can we estimate fatal parasite intensity and the host survival function?}

To compare the ability of the Adjei Method and the Likelihood Method to recover the $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above. For each parameter
combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates over pre-mortality host population sizes of $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]. We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.  Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.

\subsection*{Efficacy of the Likelihood Method with unknown pre-mortality parameters}

In the final simulation, we tested the ability of the Likelihood Method to
correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality
parameters are unknown. The previous simulations showed that the Likelihood
Method effectively identified PIHM when $\mu_p$ and $k_p$ were known with
values of 10 and 1, respectively.  As a best-case scenario, we simulated host-
parasite systems with these pre-mortality parameters and tested the power of
the Likelihood Method to identify PIHM for gradual, moderate and steep survival
functions when the pre-mortality parameters $\mu_p$ and $k_p$ also needed to be estimated.  We
perform 500 simulations over a range of different samples sizes following the
simulation procedure described above.

@@

We tested the ability of the Adjei Method and the Likelihood Method to identify
PIHM in 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
given in \cite{Adjei1986} (Table \ref{table:pihm}). \citeauthor{Crofton1971a} analyzed infection patterns in the snail \emph{Gammarus pulex} infected with the
acanthocephalan \emph{Polmorphus minutus}. \citeauthor{Adjei1986} analyzed males and females of two species of lizard fish \emph{Saurida tumbil} and
\emph{S. undosquamis} that were infected by the cestode
\emph{Callitetrarhynchus gracilis}.

In both earlier studies, the authors reported PIHM in some of the datasets and we tested whether the Adjei
Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
\cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
Method to estimate the pre-mortality distribution, and then ran the Likelihood
Method and Adjei Method using these pre-mortality parameters.  For the
\cite{Adjei1986} datasets, we followed the same procedure as the authors and
first truncated the data at 2 parasites and then fit the Crofton Method for the
female fish of both species.  Then, following the original authors' methods, we parameterized the male pre-mortality
distributions for each species with the results from the females.  Finally, we
applied the Adjei Method and the Likelihood Method to determine whether or not
PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the code provided in \emph{SI} 4.

@@

The power of the Adjei Method to detect PIHM in a
system was close to unity for larger sample sizes and tended to
decrease as sample size decreased (Figure \ref{fig:question1}C; \emph{SI} 3 Figs 1-3).  The Likelihood Method had a power close to
unity for all parameter combinations and sample sizes considered.  With gradual
survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. \ref{fig:question1}C, \emph{SI} 3 Figs 1-3).

>>>>

% \begingroup
\setlength{\parindent}{0in}
\renewcommand{\arraystretch}{2}
% \renewcommand{\tabcolsep}{0}
%\renewcommand{\tabrowsep}{1}
\begin{tabular}{p{3cm} p{10cm}}

\large\textbf{} & \vspace\baselineskip \\

\textbf{Title:} & Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations \\

\textbf{Running title:} & Methods for quantifying parasite-induced host mortality  \\

\textbf{Authors:} & Mark Q. Wilber, Sara B. Weinstein and Cheryl J. Briggs \\

\textbf{Statement of Authorship:} & MW and SW designed research, performed
analyses and wrote the manuscript, CB designed research and contributed to revisions. \\

\textbf{Affiliations:} &
1) Department of Ecology, Evolution and Marine Biology, University of California,
Santa Barbara, Santa Barbara California \\

\textbf{Email addresses:} &
Mark Wilber: mark.wilber@lifesci.ucsb.edu \newline
Sara Weinstein: sara.weinstein@lifesci.ucsb.edu \newline
Cheryl Briggs: cherie.briggs@lifesci.ucsb.edu \\

\textbf{Keywords:} & aggregation, negative binomial distribution, parasite intensity, Crofton Method, Likelihood Method
\\

\textbf{Corresponding \newline author:} &
Mark Wilber \newline
University of California, Santa Barbara \newline
Department of Ecology, Evolution, and Marine Biology \newline
2111 Noble Hall \newline
Santa Barbara, CA 93106 \newline
505-321-5381 \newline
mark.wilber@lifesci.ucsb.edu

\end{tabular}

% \renewcommand{\arraystretch}{1}
% \begin{tabular}{l l}

% \textbf{Words - Abstract:} & 293 \\
% \textbf{Words - Main Text:} & 4988 \\
% \textbf{References:} & 50 \\
% \textbf{Figures:} & 5 \\
% \textbf{Tables:} & 1

% \end{tabular}

\abstract{

Parasites can significantly impact animal populations by changing host
behavior, reproduction and survival. Detecting and quantifying these impacts is
critical for understanding disease dynamics and managing wild animal
populations. However, for wild hosts infected with macroparasites, it is
notoriously difficult to quantify the fatal parasite load and number of animals
that have died due to disease. When ethical or logistical constraints prohibit
experimental determination of these values, examination of parasite intensity
and distribution data may offer an alternative solution. In this study we
introduce a novel method for using intensity data to detect and quantify
parasite-induced mortality in wildlife populations. Using simulations, we show
that this method is more reliable than previously proposed methods while
providing quantitative estimates of parasite-induced mortality from empirical
data that are consistent with previously published qualitative estimates.
However, we stress that this method, and all techniques that estimate parasite-
induced mortality from intensity data alone, have several critical assumptions
that limit their applicability in the real world. Due to these limitations,
these methods should only be used as an exploratory tool to inform more
rigorous studies of parasite-induced host mortality.

\textbf{Keywords}: parasite aggregation, negative binomial distribution

}

@@

Infectious agents can impact on animal populations through changing
population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
of these infectious agents in wildlife is critical to understanding what
regulates host and parasite populations, making predictions about disease
transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of microparasite pathogens, such as rabies \citep{Coyne1989}, bovine tuberculosis \citep{Cox2005}, and
rinderpest \citep{Tille1991}, are typically modeled based on the presence or absence of disease, and host survival is not a function of the number of infectious agents present.  In contrast, for macroparasites, pathology increases with parasite burden and host survival probability must be treated as a function of infection intensity. Helminths exhibiting this intensity-dependent pathology have significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013}, and wildlife survival \citep{Kirk2003,Logiudice2003}. While it is generally assumed that some fraction of wild host populations succumb to parasitic infection, it is notoriously difficult to actually quantify parasite-induced host mortality (PIHM) in wild animal populations because it is difficult to observe the dead or dying hosts most impacted by parasitism \citep{McCallum2000a}.

Ideally, parasite-induced host mortality is
quantified by experimentally infecting and tracking individual hosts in the
wild population; however, for logistical and ethical reasons this method is
rarely feasible \citep{McCallum2000a}. Snapshot data of parasite intensities across multiple hosts is much easier to collect and has
often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and to quantify the
relationship between infection intensity and host mortality \citep{Adjei1986}.

\cite{Crofton1971a} first proposed that PIHM could be identified from parasite intensity data by comparing the
observed parasite distribution in sampled hosts to the distribution
predicted in the absence of parasite-induced mortality. This method
assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts. Assuming mortality occurs only in heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
fitting a negative binomial distribution to hosts with lower and lower parasite intensities, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data (Figure \ref{fig:crofton}, see \emph{Supplementary Material (SI)} 1 for additional detail).

While the Crofton Method detects the presence of PIHM, it makes no attempt to
quantify the relationship between infection intensity and host survival
probability; an important relationship for determining the potential impact of
a parasite on host dynamics \citep{AndersonandMay1978s}. \cite{Adjei1986}
suggested that this relationship could be calculated by first using the Crofton
Method to estimate the pre-mortality parasite distribution and then using this
distribution to calculate the probability of host survival with increasing
parasite intensity. To do this, \cite{Adjei1986} modeled host survival as a
logistic function and then used a generalized linear model (GLM) to estimate
the parameters of the host survival function (see \emph{SI} 2 for a technical
description of the Adjei Method). \citeauthor{Adjei1986} noted that this
survival function could provide useful information such as the parasite
intensity at which a host has a greater than 50\% chance of suffering parasite-
induced mortality. However, if mean infection intensity is high or sample sizes
are small the observed intensity data must be subjectively binned into
intensity ranges in order to fit the GLM framework. Furthermore, for the Adjei
Method to work, any observed intensity values greater than predicted values
must be modified and set equal to the predict values (see \emph{SI} 2 for
details); a questionable act of data manipulation. These manipulations may introduce bias, reduce the precision and
limit the power of this method to detect and quantify parasite-induced host
mortality.

After 30 years, and despite clear limitations \citep{McCallum2000a}, these
methods (particularly the Crofton Method) are still discussed among
parasitologists and are the primary techniques for examining population-level
impacts of parasitism using parasite intensity data. In these methods, PIHM can only be identified by visually examining plots of the pre-mortality parameters predicted by Crofton Method and determining whether they show a ``kink'' over a range of truncation values \citep[Figure \ref{fig:crofton}B;][]{Lester1984,Ferguson2011}. This qualitative criteria makes it difficult to compare PIHM between studies and a more rigorous and quantitative method is needed to both detect and quantify host mortality. The survival function given by the Adjei
Method may be used to do this; however, it requires manipulating the
original data and its accuracy remains untested.

In this study, we propose a novel method for detecting and quantifying PIHM that ameliorates many of the aforementioned deficiencies of the previous methods. Our method does not require data alteration, is highly generalizable, and uses standard statistical techniques to quantitatively determine whether PIHM is occurring in a system.  We
use simulations to compare our method with the Adjei
Method to test the ability of both to (1) detect occurrence of PIHM and (2) estimate the host survival function.  We then
apply both methods to real datasets previously used in PIHM analyses and
compare the results. Finally, we discuss the limitations of inferring PIHM
from intensity data and the place of any of these methods in modern quantitative parasitology.

@@

Our method (henceforth the Likelihood Method) begins with the same assumptions as the Adjei Method: namely that infection of
a host, parasite-induced mortality of a host, and the sampling of a host
population occur at distinct time intervals during a host's life. As discussed by
\citeauthor{Adjei1986}, this is not necessarily an unrealistic as some parasite infections are host age dependent with mortality occurring soon after infection \citep{Schotthoefer2003,Johnson2008}. [CHAT WITH SARA ABOUT THIS]

The Likelihood Method then assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \boldsymbol{\phi})$, which specifies the probability of a host having $x$ parasites when it is observed.  $\boldsymbol{\phi}$ is a vector of parameters that describes the shape of this distribution. The method then assumes that the probability of a host surviving with $x$ parasites from infection until sampling is given by the host survival function $h(\text{survival} ; x, \boldsymbol{\theta})$ where $\boldsymbol{\theta}$ specifies any additional parameters needed to define the host survival function.

With these two assumptions, we can define a distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival, $P(x | \text{survival})$.  Using standard rules of conditional probability this distribution can be written as

@@

$P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \boldsymbol{\theta})$, $P(x)$ is the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})$. Therefore, equation \ref{eq:concept} can be written as

\begin{equation}
    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}{\sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}
    \label{eq:dist}
\end{equation}

Using this probability distribution, one can then find the parameters $\boldsymbol{\theta}$
and $\boldsymbol{\phi}$ that maximize the likelihood of an observed host-parasite dataset.
To estimate the significance of PIHM in a host-parasite system, a likelihood
ratio test can be used in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by the pre-mortality distribution
$g(x; \boldsymbol{\phi})$.  If PIHM is not significant in the system, the resulting
likelihood ratio statistic should approximately follow a $\chi^2$ distribution
with degrees of freedom equal to the number of parameters in the full model with parasite-induced mortality minus the number of parameters in the reduced model without parasite-induced mortality \citep{Bolker2008}.

The parameterization of equation \ref{eq:dist} depends on the parasite system of interest.  Here, we assume that the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ follows a negative binomial
distribution with the parameters mean parasite intensity ($\mu_p$) and
aggregation ($k_p$) before mortality (smaller $k_p$ indicates
more aggregation) \citep{Crofton1971a,AndersonandMay1978,Adjei1986}. A variety of different biological and statistical assumptions can result in an equilibrium parasite distribution that follows a negative binomial distribution
\citep{Kendall1948a, Boswell1970, Calabrese2011}. Furthermore, the negative binomial distribution is an incredibly flexible distribution that fits many host-parasite systems even when the underlying mechanisms determining the empirical distribution are unknown \citep{Shaw1998}.

The function for $h(\text{survival}; x,
\boldsymbol{\theta})$ is also system specific.  Many
theoretical models of parasite-induced host mortality assume that the parasite-induced death rate of hosts is a linear function of parasite intensity
\citep{AndersonandMay1978,Dobson1992,Barbour2000}. In systems where there is truly a linear relationship between infection intensity and survival probability it will be nearly impossible to use intensity data to detect parasite-induced host mortality \citep{Lanciani1989}.  However,  there is substantial empirical evidence that host-parasite systems do not universally exhibit linear host-survival functions \citep{Benesh2011}, suggesting a linear form for $h(\text{survival} ; x, \boldsymbol{\theta}$) is not always the most reasonable.

To compare the Likelihood Method and the previously proposed Adjei Method, we adopt the host-survival function used in the earlier study and assume it is non-linear and follows a logistic function given by

\begin{equation}
    h(\text{survival}; x, a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}

Generally, a larger $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases, with the maximum rate of decline having a value of $b / 4$ (\emph{SI} 2). $b$ is in many ways analogous to the the pathogenicity parameter in classic macroparasite models which gives that rate at which host death rate changes with increasing parasite intensity \citep{AndersonandMay1978,Isham1995}. When $b$ is held constant, a larger $a$ allows for hosts to tolerate larger parasite intensities before experiencing parasite-induced mortality. More specifically, for every one unit increase in $a$ the log parasite intensity at which any percent of hosts survive (e.g. 99\% of hosts survive) increases by $ 1 / b$ (\emph{SI} 2).

The equation $\exp(a
/ b)$ can also be used to calculate the parasite $LD_{50}$, here defined as the
infection intensity above which a host has greater than 50\% probability of dying.  Equation \ref{eq:logistic} is commonly used in toxicology and has the useful properties of being bounded between 0 and 1 and being differentiable for all $x$ \citep{Collet2002}.  That being said, it is phenomenological and there is little theoretical justification to use it rather than it tends to fit survival data. However, given that a goal of these analyses is to
compare the Likelihood Method's results to the Adjei Method, it is natural
to adopt the same host-survival function to facilitate comparison.  When
applying the Likelihood Method to other systems more mechanistic host-survival functions can be used in place of equation \ref{eq:logistic}.

@@

We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters. Consistent with the model assumption that parasite infection, host mortality, and population sampling are temporally separate events, we first created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$. This represents a host population that has become infected but not yet experienced parasite-induced mortality \citep{Adjei1986}.  In the Adjei Method and Crofton Method, $N_p$ is a necessary parameter defined as the number of hosts in the population before parasite-induced mortality. More accurately, $N_p$  is the number of hosts that would have been sampled had parasite-induced host mortality not occurred.  This parameter is not necessary when using the Likelihood Method because, unlike the Adjei Method and Crofton Method which estimate parasite-induced mortality using absolute numbers of hosts, the Likelihood Method estimates parasite-induced mortality using probabilities. However, to compare the results of the Likelihood Method with the Adjei Method, we specified a value for $N_p$ for all simulations.

We next chose values of $a$ and $b$ for the host survival function and calculated the probability of survival
for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, to simulate the period in which hosts died due to infection, for each host we drew a random number from a uniform distribution
between 0 and 1 and if the calculated host survival probability was less than this random
number, the host experienced parasite-induced mortality.  The surviving individuals represent the post-mortality hosts that would be sampled in the field.

We then used these simulated pre-mortality and post-mortality datasets to test the
ability of both methods to correctly determine whether or not PIHM was
occurring when the parameters $N_p$, $\mu_p$ and $k_p$ were known.  Although
the parameters $N_p$, $\mu_p$, and $k_p$ are always unknown in real systems, a
method that fails under these ideal simulation conditions with known parameters will certainly also fail when these values must be estimated from empirical data. In practice, for the Adjei Method, $N_p$, $\mu_p$,
and $k_p$ are estimated using the Crofton Method \citep{Adjei1986}, while $\mu_p$ and $k_p$ in
the Likelihood Method can be estimated jointly with $a$ and $b$ or via the
Crofton Method.

We compared the two methods using three different mean parasite intensity values ($\mu_p$ = 10, 50, 100) and three different host survival functions (gradual, moderate, and steep decreases in the host survival with increasing parasite intensity, Figure \ref{fig:question1}A). For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these 27 parameter
combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.  We also examined the impact of sample size by simulating each parameter for pre-mortality sample sizes of $N_p$ = [50, 100, 200, 300, 400, 500].  Wild host populations were assumed to be sampled after PIHM has occurred, thus we calculated the sample size in the power simulations as the average number of surviving hosts over all 150 simulations for each parameter combination.

We then tested the ability of the Likelihood Method to correctly identify PIHM
under the more realistic condition of unknown pre-mortality parameters. Based
on the first set of simulations, we excluded the Adjei Method and only examined
the power of the Likelihood Method under ``best-case'' scenario parameter
values, setting $\mu_p = 10$ and $k = 1$ because PIHM is most detectable when
parasites are less clumped and mean intensity is low. We examined the impact of survival function shape and
sample size on the Likelihood Method's ability to identify PIHM when the pre-mortality parameters $\mu_p$ and $k_p$ and the survival function parameters $a$
and $b$ needed to be estimated.  We performed 500 simulations over a range of
different samples sizes for gradual, moderate, and steep survival functions, following the simulation procedure described above. \\

\noindent
\emph{Question 2: Can we estimate properties of the host survival function?}

In the previous section we compared the ability of the Adjei Method and the Likelihood Method to provide a ``yes'' or ``no'' answer for whether or not PIHM was occurring in a system. In this section we compare the ability of the Adjei Method and the Likelihood Method to estimate properties of the survival function such as the parameters $a$, $b$ and $LD_{50}$.  Using the same simulation procedure and parameter combinations described above, we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision for these estimates \citep{Walther2005}.  Because estimating properties of the host survival function requires more information than simply detecting PIHM, we used larger values of $N_p$ for this simulation ($N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]).  We used the average number of surviving hosts for each set of 150 simulated datasets as our measure of sample size.  Although both $a$ and $b$ are necessary to estimate $LD_{50}$, the two parameters showed similar patterns of bias and precision so we only show the results for $a$.

@@

We tested the ability of the Adjei Method and the Likelihood Method to identify
PIHM in six host-parasite datasets given in \cite{Crofton1971a} and four datasets
given in \cite{Adjei1986} (Table \ref{table:pihm}). \citeauthor{Crofton1971a} analyzed infection patterns in the snail \emph{Gammarus pulex} infected with the
acanthocephalan \emph{Polmorphus minutus}. \citeauthor{Adjei1986} analyzed males and females of two species of lizard fish \emph{Saurida tumbil} and
\emph{S. undosquamis} that were infected by the cestode
\emph{Callitetrarhynchus gracilis}.

In both earlier studies, the authors reported PIHM in some of the datasets and we tested whether the Adjei
Method and/or the Likelihood Method also predicted PIHM. For the six datasets from
\cite{Crofton1971a}, we used the general conclusions of the author and truncated the data at four parasites, applied the Crofton
Method to estimate the pre-mortality distribution, and then ran the Likelihood
Method and Adjei Method using these pre-mortality parameters.  For the
\cite{Adjei1986} datasets, we followed the same procedure as the authors and
first truncated the data at two parasites and then fit the Crofton Method for the
female fish of both species.  Then, following the original authors' methods, we parameterized the male pre-mortality
distributions for each species with the results from the females.  Finally, we
applied the Adjei Method and the Likelihood Method to determine whether or not
PIHM was significant for these species and compared our results to those given by the authors.  All code for the analyses is provided in \emph{SI} 4.

@@

The power of the Adjei Method to detect PIHM in a
system was close to unity for larger sample sizes and tended to
decrease as sample size decreased for all survival functions (Figure \ref{fig:question1}C; \emph{SI} 3 Figs 1-3).  The Likelihood Method had a power close to
unity for all parameter combinations and sample sizes considered.  With gradual
survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. \ref{fig:question1}C, \emph{SI} 3 Figs 1-3).

The Adjei Method showed highly inflated Type I error rates (i.e. falsely detected
PIHM) for all parameter combinations that we
considered (Fig. \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).  This method also showed the unintuitive pattern of decreasing Type I error
rate with decreasing sample size.  This occurred because, at small samples sizes, intensity data must be binned before the Adjei Method can be used (\emph{SI} 2).  In contrast, the Likelihood Method showed a Type I
error rate at or near the pre-set level of 0.05 for all parameter combinations
and sample sizes considered (Fig. \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).

When all parameters were jointly estimated, the Likelihood Method showed highly
context-dependent results even when detecting PIHM under the best-case scenario of
$\mu_p = 10$ and $k_p = 1$. For steep survival curves, PIHM could be detected with a power of greater than 0.8 from a sample of less than 100 hosts (Fig \ref{fig:real_power}). However, for moderate survival functions over 400 hosts had to be sampled to achieve the same power and for gradual survival functions, no tested sample size ever achieved a power greater than 0.8 (Fig \ref{fig:real_power}).
